---
title: "Inspect Grit"
author: "Shantanu Singh"
output: 
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: lumen
---

```
output: 
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: lumen
```

```
output: github_document
```

# Boilerplate 

```{r echo=FALSE}
show_table <- print
```

If running interactively in RStudio, 

- change `output` in the header of this markdown to `html_notebook` and
- change to `eval=TRUE` below

When knitting for pushing to GitHub,

- change `output` in the header of this markdown to `github_document` and
- change to `eval=FALSE` below

```{r eval=FALSE}
show_table <- knitr::kable
knitr::opts_chunk$set(fig.width = 6, fig.asp = 2/3)
```


```{r message=FALSE}
library(magrittr)
library(tidyverse)
library(arrow)

```


```{r}
source("../../R/cytominer-eval.R")
```

# Load data

Load EMPTY-normalized data

```{r message=FALSE}
profiles <- read_csv("data/cell_health_merged_feature_select.csv.gz")
```


```{r}
metadata <- get_annotation(profiles)
```


```{r}
head(metadata, 5)
```


# Calculate similarity

```{r}
if (file.exists("data/sim.parquet")) {
  message("Loading existing similarity file...")
  sim_df <- arrow::read_parquet("data/sim.parquet")
    
} else {
  sim_df <- sim_calculate(profiles)
  sim_df %>% arrow::write_parquet("data/sim.parquet")
  
}
```

# Replicate similarity

## Calculate

```{r}
all_same_cols_rep <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name")

annotation_cols <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name")

rep <-
  sim_all_same(sim_df,
               metadata,
               all_same_cols_rep,
               annotation_cols,
               drop_lower = TRUE)

all_same_cols_rep_ref <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name",
    "Metadata_Plate")

```

## Check

```{r}
n_profiles <- 
  profiles %>%
  group_by(across(all_of(all_same_cols_rep))) %>%
  tally(name = "n_profiles") %>%
  arrange(desc(n_profiles))

n_replicate_pairs <- 
  rep %>%
  group_by(across(all_of(annotation_cols))) %>%
  tally(name = "n_replicate_pairs") %>%
  arrange(desc(n_replicate_pairs))

inner_join(n_profiles, n_replicate_pairs, by = all_same_cols_rep) %>%
  mutate(check =
           n_replicate_pairs == (n_profiles * (n_profiles - 1)) / 2) %>%
  filter(!check)

```

# Reference similarity

## Calculate

```{r}
filter_keep_right <-
  data.frame(Metadata_gene_name = "EMPTY")

annotation_cols <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name")

all_same_cols_ref <-
  c("Metadata_cell_line",
    "Metadata_Plate")

annotation_cols <-
  unique(c(annotation_cols, all_same_cols_ref))

ref <-
  sim_all_same_keep_some(sim_df,
                         metadata,
                         all_same_cols_ref,
                         filter_keep_right,
                         annotation_cols)
```

## Check 

```{r}
all_same_cols_rep <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name")

all_same_cols_ref <-
  c("Metadata_cell_line", 
    "Metadata_Plate")

# number of profiles
n_profiles <- 
  profiles %>%
  group_by(across(all_of(c(all_same_cols_rep, all_same_cols_ref)))) %>%
  tally(name = "n_profiles") %>%
  arrange(desc(n_profiles))

# number of references
n_references <-
  metadata %>%
  inner_join(filter_keep_right, by = colnames(filter_keep_right)) %>%
  group_by(across(all_of(all_same_cols_ref))) %>%
  tally(name = "n_references")

# number of pairs involving the reference (on the right column)
n_reference_pairs <- 
  ref %>%
  group_by(across(all_of(c(all_same_cols_rep, all_same_cols_ref)))) %>%
  tally(name = "n_reference_pairs") %>%
  arrange(desc(n_reference_pairs))

# number of profiles x number of references = number of pairs involving the
# reference (on the right column)
check_df <- 
  n_references %>% 
  inner_join(n_reference_pairs, by = all_same_cols_ref) %>%
  inner_join(n_profiles, by = unique(c(all_same_cols_rep, all_same_cols_ref)))

check_df %>%
  mutate(check =
           n_reference_pairs == (n_profiles * n_references)) %>%
  filter(!check)
```


# Non-replicate similarity

## Calculate

```{r}

all_same_cols_rep <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name")

all_same_cols_ref <-
  c("Metadata_cell_line", 
    "Metadata_Plate")


any_different_cols_non_rep <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name")

all_same_cols_non_rep <-
  c("Metadata_cell_line", "Metadata_Plate")

all_different_cols_non_rep <-
  c("Metadata_gene_name")

annotation_cols <-
  c(
    "Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name",
    "Metadata_Plate",
    "Metadata_Well"
  )

filter_keep_left <-
  data.frame(Metadata_gene_name = "EMPTY")

filter_keep_right <-
  data.frame(Metadata_gene_name = "EMPTY")

nonrep <-
  sim_some_different_keep_some(
    sim_df,
    metadata,
    any_different_cols = any_different_cols_non_rep,
    all_same_cols = all_same_cols_non_rep,
    all_different_cols = all_different_cols_non_rep,
    filter_keep_left = filter_keep_left,
    filter_keep_right = filter_keep_right
  )
```

## Check

```{r}
nonrep_annotated <- 
  nonrep %>%
  sim_annotate(metadata, annotation_cols, index = "both")

for (x in all_same_cols_non_rep) {
  x1 <- paste0(x, "1")
  x2 <- paste0(x, "2")
  # by definition of all_same_cols
  stopifnot(nonrep_annotated %>% filter(.data[[x1]] != .data[[x2]]) %>% nrow() == 0)
}

for (x in all_different_cols_non_rep) {
  x1 <- paste0(x, "1")
  x2 <- paste0(x, "2")
  # by definition of all_different_cols
  stopifnot(nonrep_annotated %>% filter(.data[[x1]] == .data[[x2]]) %>% nrow() == 0)
}

# by definition of any_different_cols
# any_different is identical to not_all_same
stopifnot(
  nonrep_annotated %>%
    unite(
      "any_different_col1",
      all_of(paste0(any_different_cols_non_rep, "1")),
      sep = ":",
      remove = FALSE
    ) %>%
    unite(
      "any_different_col2",
      all_of(paste0(any_different_cols_non_rep, "2")),
      sep = ":",
      remove = FALSE
    ) %>%
    filter(any_different_col1 == any_different_col2) %>%
    nrow() == 0
)
```


# Inspect

## Similarity among replicates

```{r}
rep %>%
  filter(Metadata_gene_name != "EMPTY") %>%
  ggplot(aes(sim)) + 
  geom_histogram(binwidth = .1) + 
  facet_wrap(~Metadata_cell_line) +
  ggtitle("Similarity of perturbations to each other", subtitle = "(EMPTY are dropped)")

```

## Similarity among reference wells


```{r}
rep %>% 
  filter(Metadata_gene_name == "EMPTY") %>%
  ggplot(aes(Metadata_cell_line, sim)) + 
  geom_boxplot() +
  ggtitle("Similarity of EMPTY wells")

rep %>% 
  filter(Metadata_gene_name == "EMPTY") %>%
  group_by(Metadata_cell_line) %>%
  summarise(across(all_of("sim"), list(mean = mean, sd = sd)))
```

## Similarity between perturbation wells and reference wells

```{r}
ref %>%
  ggplot(aes(sim)) + 
  geom_histogram(binwidth = .1) + 
  facet_wrap(~Metadata_cell_line) +
  ggtitle("Similarity of EMPTY wells")
```

```{r}
annotation_cols <- c("Metadata_cell_line", "Metadata_gene_name", "Metadata_pert_name")

ref %>% 
  group_by(across(all_of(annotation_cols))) %>%
  summarise(across(all_of("sim"), list(mean = mean, sd = sd))) %>%
  pivot_longer(-any_of(annotation_cols)) %>%
  ggplot(aes(Metadata_cell_line, value)) + geom_boxplot() + 
  facet_wrap(~name) +
  ggtitle("Similarity of perturbations to EMPTY")

```
## Compare similarities 

```{r}
# summarize similarity to replicates
sum_rep <-
  rep %>% 
  group_by(across(all_of(all_same_cols_rep))) %>%
  summarise(sim_rep_q75 = quantile(sim, .75), .groups = "keep") %>%
  ungroup()

# summarize similarity to reference wells
sum_ref <-
  ref %>% 
  group_by(across(all_of(all_same_cols_rep))) %>%
  summarise(sim_ref_q75 = quantile(sim, .75), .groups = "keep")
  
sum_all <-
  sum_ref %>%
  inner_join(sum_rep)
```


```{r}
ggplot(sum_all, aes(sim_ref_q75, sim_rep_q75)) + 
  geom_point() + 
  facet_wrap(~Metadata_cell_line) +
  xlab("q75 similarity to EMPTY") +
  ylab("q75 pairwise replicate similarity") + coord_equal()
```

## Inspect non-replicates

```{r}
all_same_cols_rep <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name")

tally_cols <- c(all_same_cols_rep, "Metadata_Plate", "Metadata_Well")
tally_cols <- paste0(tally_cols, "1")

nonrep_annotated %>%
  group_by(across(all_of(tally_cols))) %>%
  tally() %>%
  head()
```

