---
title: "Inspect Grit"
author: "Shantanu Singh"
output: 
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: lumen
---

```
output: 
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: lumen
```

```
output: github_document
```

# Boilerplate 

```{r echo=FALSE}
show_table <- print
```

If running interactively in RStudio, 

- change `output` in the header of this markdown to `html_notebook` and
- change to `eval=TRUE` below

When knitting for pushing to GitHub,

- change `output` in the header of this markdown to `github_document` and
- change to `eval=FALSE` below

```{r eval=FALSE}
show_table <- knitr::kable
knitr::opts_chunk$set(fig.width = 6, fig.asp = 2/3)
```


```{r message=FALSE}
library(magrittr)
library(tidyverse)
library(arrow)
```


```{r}
source("../../R/cytominer-eval.R")
```


# Load data

Load EMPTY-normalized data

```{r message=FALSE}
profiles <- read_csv("data/cell_health_merged_feature_select.csv.gz")
```


```{r}
metadata <- get_annotation(profiles)
```


```{r}
head(metadata, 5)
```



# Calculate similarity

```{r}
if (file.exists("data/sim.parquet")) {
  message("Loading existing similarity file...")
  sim_df <- arrow::read_parquet("data/sim.parquet")
    
} else {
  sim_df <- sim_calculate(profiles)
  sim_df %>% arrow::write_parquet("data/sim.parquet")
  
}
```

# Calculate 

```{r}
annotation_cols <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name")
```

## Drop some wells

```{r}
drop_group <-
  data.frame(Metadata_gene_name = "EMPTY")

if (!is.null(drop_group)) {
  sim_df %<>%
    sim_drop(metadata, drop_group, "left") %>%
    sim_drop(metadata, drop_group, "right")
}
```

## Similarity to reference

Fetch similarities between

a. all rows (except, optionally those containing `reference`)

and 

b. all rows containing `reference`

Do so only for those (a, b) pairs that 

- have *same* values in *all* columns of `all_same_cols_ref`

```{r}
reference <-
  data.frame(Metadata_gene_name = c("Chr2", "Luc", "LacZ"))

drop_reference <- FALSE 

all_same_cols_ref <-
  c("Metadata_cell_line",
    "Metadata_Plate")

ref <-
  sim_df %>%
  sim_all_same_keep_some(
    metadata,
    all_same_cols_ref,
    filter_keep_right = reference,
    drop_reference = drop_reference,
    annotation_cols =
      unique(c(annotation_cols, all_same_cols_ref))
  )
```

## Similarity to group

Fetch similarities between

a. all rows 

and 

b. all rows

Do so only for those (a, b) pairs that 

- have *same* values in *all* columns of `all_same_cols_group`

- have *different* values in *at least one* column of `any_different_cols_group`

```{r}
all_same_cols_group <-
  c("Metadata_cell_line",
    "Metadata_gene_name")

any_different_cols_group <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name")

rep_group <-
  sim_df %>%
  sim_some_different_drop_some(
    metadata,
    any_different_cols = any_different_cols_group,
    all_same_cols = all_same_cols_group)
```


## Similarity to replicates (no references)

Fetch similarities between

a. all rows except `reference` rows

and

b. all rows except `reference` rows (i.e. to each other)

Do so for only those (a, b) pairs that 

- have *same* values in *all* columns of `all_same_cols_rep

Keep, both, (a, b) and (b, a)
    
```{r}
all_same_cols_rep <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name")

rep <-
  sim_df %>%
  sim_drop(metadata, reference, "left") %>%
  sim_drop(metadata, reference, "right") %>%
  sim_all_same(metadata,
               all_same_cols_rep,
               annotation_cols,
               drop_lower = FALSE)
```


## Similarity to replicates (only references)

Fetch similarities between

a. all rows containing `reference`

and

b. all rows containing `reference` (i.e. to each other)

Do so for only those (a, b) pairs that 

- have *same* values in *all* columns of `all_same_cols_rep_ref`. 

Keep, both, (a, b) and (b, a)

```{r}
all_same_cols_rep_ref <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name",
    "Metadata_Plate")

rep_ref <-
  sim_df %>%
  sim_keep(metadata,
           reference,
           "left") %>%
  sim_keep(metadata,
           reference,
           "right") %>%
  sim_all_same(
    metadata,
    all_same_cols_rep_ref,
    annotation_cols = all_same_cols_rep_ref,
    drop_lower = FALSE
  )
```


## Similarity to non-replicates

Fetch similarities between

a. all rows (except, optionally, `reference` rows)

and

b. all rows except `reference` rows

Do so for only those (a, b) pairs that 

- have *same* values in *all* columns of `all_same_cols_non_rep`

- have *different* values in *all* columns `all_different_cols_non_rep`

- have *different* values in *at least one* column of `any_different_cols_non_rep`

Keep, both, (a, b) and (b, a)

```{r}
any_different_cols_non_rep <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name")

all_same_cols_non_rep <-
  c("Metadata_cell_line", "Metadata_Plate")

all_different_cols_non_rep <-
  c("Metadata_gene_name")


if (drop_reference) {
  reference_left <- reference
} else {
  reference_left <- NULL
}

nonrep <-
  sim_df %>%
  sim_some_different_drop_some(
    metadata,
    any_different_cols = any_different_cols_non_rep,
    all_same_cols = all_same_cols_non_rep,
    all_different_cols = all_different_cols_non_rep,
    filter_drop_left = reference_left,
    filter_drop_right = reference
  )
```


# Check 

## Similarity to reference

```{r}
# number of profiles
n_profiles <- 
  profiles %>%
  group_by(across(all_of(c(all_same_cols_rep, all_same_cols_ref)))) %>%
  tally(name = "n_profiles") %>%
  arrange(desc(n_profiles))

if(drop_reference) {
  n_profiles %<>% anti_join(reference)
}

# number of references
n_references <-
  metadata %>%
  inner_join(reference, by = colnames(reference)) %>%
  group_by(across(all_of(all_same_cols_ref))) %>%
  tally(name = "n_references")

# number of pairs involving the reference (on the right column)
n_reference_pairs <- 
  ref %>%
  group_by(across(all_of(c(all_same_cols_rep, all_same_cols_ref)))) %>%
  tally(name = "n_reference_pairs") %>%
  arrange(desc(n_reference_pairs))

# number of profiles x number of references = number of pairs involving the
# reference (on the right column)
check_df <- 
  n_references %>% 
  inner_join(n_reference_pairs, by = all_same_cols_ref) %>%
  inner_join(n_profiles, by = unique(c(all_same_cols_rep, all_same_cols_ref)))

stopifnot(check_df %>%
            mutate(check =
                     n_reference_pairs == (n_profiles * n_references)) %>%
            anti_join(reference) %>%
            filter(!check) %>%
            nrow() == 0)

stopifnot(check_df %>%
            mutate(check =
                     n_reference_pairs == (n_profiles * n_references) - n_profiles) %>%
            inner_join(reference) %>%
            filter(!check) %>%
            nrow() == 0)
```

## Similarity to replicates (no references)

```{r}
n_profiles <- 
  profiles %>%
  anti_join(reference, by = colnames(reference)) %>%
  group_by(across(all_of(all_same_cols_rep))) %>%
  tally(name = "n_profiles") %>%
  arrange(desc(n_profiles))

n_replicate_pairs <- 
  rep %>%
  group_by(across(all_of(annotation_cols))) %>%
  tally(name = "n_replicate_pairs") %>%
  arrange(desc(n_replicate_pairs))

stopifnot(
  inner_join(n_profiles, n_replicate_pairs, by = all_same_cols_rep) %>%
    mutate(check =
             n_replicate_pairs == n_profiles * (n_profiles - 1)) %>%
    filter(!check) %>%
    nrow() == 0
)

```

## Similarity to nonreplicates

```{r}
nonrep_annotated <-
  nonrep %>%
  sim_annotate(
    metadata,
    annotation_cols = c(annotation_cols, "Metadata_Plate", "Metadata_Well"),
    index = "both"
  )

walk(all_same_cols_non_rep,
     function(x) {
       x1 <- paste0(x, "1")
       x2 <- paste0(x, "2")
       # by definition of all_same_cols
       stopifnot(nonrep_annotated %>% filter(.data[[x1]] != .data[[x2]]) %>% nrow() == 0)
     })


walk(all_different_cols_non_rep,
     function(x) {
       x1 <- paste0(x, "1")
       x2 <- paste0(x, "2")
       # by definition of all_different_cols
       stopifnot(nonrep_annotated %>% filter(.data[[x1]] == .data[[x2]]) %>% nrow() == 0)
     })

# by definition of any_different_cols
# any_different is identical to not_all_same
stopifnot(
  nonrep_annotated %>%
    unite(
      "any_different_col1",
      all_of(paste0(any_different_cols_non_rep, "1")),
      sep = ":",
      remove = FALSE
    ) %>%
    unite(
      "any_different_col2",
      all_of(paste0(any_different_cols_non_rep, "2")),
      sep = ":",
      remove = FALSE
    ) %>%
    filter(any_different_col1 == any_different_col2) %>%
    nrow() == 0
)
```

## Similarity to group

```{r}
repgroup_annotated <-
  rep_group %>%
  sim_annotate(
    metadata,
    annotation_cols = annotation_cols,
    index = "both"
  )

walk(all_same_cols_group,
     function(x) {
       x1 <- paste0(x, "1")
       x2 <- paste0(x, "2")
       # by definition of all_same_cols
       stopifnot(repgroup_annotated %>% filter(.data[[x1]] != .data[[x2]]) %>% nrow() == 0)
     })

# by definition of any_different_cols
# any_different is identical to not_all_same
stopifnot(
  repgroup_annotated %>%
    unite(
      "any_different_col1",
      all_of(paste0(any_different_cols_group, "1")),
      sep = ":",
      remove = FALSE
    ) %>%
    unite(
      "any_different_col2",
      all_of(paste0(any_different_cols_group, "2")),
      sep = ":",
      remove = FALSE
    ) %>%
    filter(any_different_col1 == any_different_col2) %>%
    nrow() == 0
)
```


```{r}
repgroup_annotated %>%
  group_by(across(all_of(paste0(all_same_cols_group, "1")))) %>%
  tally()
```

# Inspect

## Similarity to replicates (no references)

```{r}
rep %>%
  ggplot(aes(sim)) + 
  geom_histogram(binwidth = .1) + 
  facet_wrap(~Metadata_cell_line) +
  ggtitle("Similarity among replicates")
```
## Similarity to group

```{r}
repgroup_annotated <-
  rep_group %>%
  sim_annotate(
    metadata,
    annotation_cols = annotation_cols,
    index = "left"
  )

repgroup_annotated %>%
  anti_join(reference, by = colnames(reference)) %>%
  ggplot(aes(sim)) + 
  geom_histogram(binwidth = .1) + 
  facet_wrap(~Metadata_cell_line) +
  ggtitle("Similarity to group (exclude references)")
```
```{r}
repgroup_annotated %>%
  inner_join(reference, by = colnames(reference)) %>%
  ggplot(aes(sim)) + 
  geom_histogram(binwidth = .1) + 
  facet_wrap(Metadata_gene_name~Metadata_cell_line, scales = "free_y") +
  ggtitle("Similarity to group (only references)")


repgroup_annotated %>%
  inner_join(reference, by = colnames(reference)) %>%
  ggplot(aes(Metadata_cell_line, sim, fill = Metadata_gene_name)) + 
  geom_boxplot( position = "dodge") +
  ggtitle("Similarity to group (only references)")

repgroup_annotated %>%
  inner_join(reference, by = colnames(reference)) %>%
  group_by(Metadata_cell_line) %>%
  summarise(across(all_of("sim"), list(mean = mean, sd = sd)))

```

## Similarity to replicates (only references)

```{r}
rep_ref %>%
  ggplot(aes(sim)) + 
  geom_histogram(binwidth = .1) + 
  facet_wrap(Metadata_gene_name~Metadata_cell_line, scales = "free_y") +
  ggtitle("Similarity among reference replicates")

rep_ref %>% 
  ggplot(aes(Metadata_cell_line, sim, fill = Metadata_gene_name)) + 
  geom_boxplot( position = "dodge") +
  ggtitle("Similarity among reference replicates")

rep_ref %>% 
  group_by(Metadata_cell_line) %>%
  summarise(across(all_of("sim"), list(mean = mean, sd = sd)))
```

## Similarity to reference

```{r}
ref %>%
  anti_join(reference) %>%
  ggplot(aes(sim)) + 
  geom_histogram(binwidth = .1) + 
  facet_wrap(~Metadata_cell_line) +
  ggtitle("Similarity to reference (exclude references)")
```


```{r}
ref %>%
  inner_join(reference) %>%
  ggplot(aes(sim)) + 
  geom_histogram(binwidth = .1) + 
  facet_wrap(Metadata_gene_name~Metadata_cell_line, scales = "free_y") +
  ggtitle("Similarity to reference (only references)")
```

```{r}
ref %>% 
  anti_join(reference, by = colnames(reference)) %>%
  group_by(across(all_of(annotation_cols))) %>%
  summarise(across(all_of("sim"), list(mean = mean, sd = sd)), .groups = "keep") %>%
  ungroup() %>%
  pivot_longer(-any_of(annotation_cols)) %>%
  ggplot(aes(Metadata_cell_line, value)) + geom_boxplot() + 
  facet_wrap(~name) +
  ggtitle("Similarity to reference (exclude references)", 
          subtitle = "Distributions of mean and s.d. for each perturbation")
```
```{r}
ref %>% 
  inner_join(reference, by = colnames(reference)) %>%
  group_by(across(all_of(annotation_cols))) %>%
  summarise(across(all_of("sim"), list(mean = mean, sd = sd)), .groups = "keep") %>%
  ungroup() %>%
  pivot_longer(-any_of(annotation_cols)) %>%
  ggplot(aes(Metadata_cell_line, value, fill = Metadata_gene_name)) + geom_boxplot(position = "dodge") + 
  facet_wrap(~name) +
  ggtitle("Similarity to reference (only references)", 
          subtitle = "Distributions of mean and s.d. for each perturbation")
```

## Similarity to non-replicates

```{r}
nonrep_annotated <-
  nonrep %>%
  sim_annotate(metadata,
               annotation_cols = c(annotation_cols),
               index = "left")

nonrep_annotated %>%
  anti_join(reference, by = colnames(reference)) %>%
  ggplot(aes(sim)) + 
  geom_histogram(binwidth = .1) + 
  facet_wrap(~Metadata_cell_line) +
  ggtitle("Similarity among non-replicates (exclude references)")
```


```{r}
nonrep_annotated %>%
  inner_join(reference, by = colnames(reference)) %>%
  ggplot(aes(sim)) + 
  geom_histogram(binwidth = .1) + 
  facet_wrap(Metadata_gene_name~Metadata_cell_line, scales = "free_y") +
  ggtitle("Similarity among non-replicates (include references)")
```

```{r}
nonrep_annotated %>% 
  anti_join(reference, by = colnames(reference)) %>%
  group_by(across(all_of(annotation_cols))) %>%
  summarise(across(all_of("sim"), list(mean = mean, sd = sd)), .groups = "keep") %>%
  pivot_longer(-any_of(annotation_cols)) %>%
  ggplot(aes(Metadata_cell_line, value)) + geom_boxplot() + 
  facet_wrap(~name) +
  ggtitle("Similarity among non-replicates (exclude references)", 
          subtitle = "Distributions of mean and s.d. for each perturbation")
```


```{r}
nonrep_annotated %>% 
  inner_join(reference, by = colnames(reference)) %>%
  group_by(across(all_of(annotation_cols))) %>%
  summarise(across(all_of("sim"), list(mean = mean, sd = sd)), .groups = "keep") %>%
  ungroup() %>%
  pivot_longer(-any_of(annotation_cols)) %>%
  ggplot(aes(Metadata_cell_line, value, fill = Metadata_gene_name)) + 
  geom_boxplot(position = "dodge") + 
  facet_wrap(~name) +
  ggtitle("Similarity among non-replicates (include references)", 
          subtitle = "Distributions of mean and s.d. for each perturbation")

```




