---
title: "Deconstruct Grit"
author: "Shantanu Singh"
output: 
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: lumen
---

```
output: 
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: lumen
```

```
output: github_document
```

# Boilerplate 

```{r echo=FALSE}
show_table <- print
```

If running interactively in RStudio, 

- change `output` in the header of this markdown to `html_notebook` and
- change to `eval=TRUE` below

When knitting for pushing to GitHub,

- change `output` in the header of this markdown to `github_document` and
- change to `eval=FALSE` below

```{r eval=FALSE}
show_table <- knitr::kable
knitr::opts_chunk$set(fig.width = 6, fig.asp = 2/3)
```

# Libraries

```{r message=FALSE}
library(magrittr)
library(tidyverse)
library(arrow)
```


```{r}
source("../../R/cytominer-eval.R")
```

# Load data

```{r message=FALSE}
# read these features
# https://github.com/broadinstitute/grit-benchmark/blob/31d9812e0c267d3535c85a72ef6bd0104d1a2130/1.calculate-metrics/cell-health/0.calculate-grit.ipynb
# (See chunk 4)

profiles <- read_csv("data/cell_health_merged_feature_select.csv.gz")

metadata <- get_annotation(profiles)
```


```{r}
# if (file.exists("data/sim.parquet")) {
#   message("Loading existing similarity file...")
#   sim_df <- arrow::read_parquet("data/sim.parquet")
#     
# } else {
#   sim_df <- sim_calculate(profiles)
#   sim_df %>% arrow::write_parquet("data/sim.parquet")
# }
# 
sim_df <- sim_calculate(profiles)
```

# Compute similarity sets

```{r}

# ---- 0. Filter out some rows ----

drop_group <-
  data.frame(Metadata_gene_name = "EMPTY")

# ---- 1. Similarity to reference ----

# Fetch similarities between
# a. all rows (except, optionally those containing `reference`)
# and
# b. all rows containing `reference`
# Do so only for those (a, b) pairs that
# - have *same* values in *all* columns of `all_same_cols_ref`
    
reference <-
  data.frame(Metadata_gene_name = c("Chr2", "Luc", "LacZ"))

# all_same_cols_ref <-
#   c("Metadata_cell_line")

all_same_cols_ref <-
  c("Metadata_cell_line",
    "Metadata_Plate")

# ---- 2. Similarity to replicates (no references) ----

# Fetch similarities between
# a. all rows except `reference` rows
# and
# b. all rows except `reference` rows (i.e. to each other)
#
# Do so for only those (a, b) pairs that
# - have *same* values in *all* columns of `all_same_cols_rep
#
# Keep, both, (a, b) and (b, a)

all_same_cols_rep <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name")

# ---- 3. Similarity to replicates (only references) ----

# Fetch similarities between
# a. all rows containing `reference`
# to
# b. all rows containing `reference` (i.e. to each other)
#
# Do so for only those (a, b) pairs that
# - have *same* values in *all* columns of `all_same_cols_rep_ref`.
#
# Keep, both, (a, b) and (b, a)

all_same_cols_rep_ref <-
  c(
    "Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name",
    "Metadata_Plate"
  )

# ---- 4. Similarity to non-replicates ----

# Fetch similarities between
# a. all rows (except, optionally, `reference` rows)
# to
# b. all rows except `reference` rows
#
# Do so for only those (a, b) pairs that
# - have *same* values in *all* columns of `all_same_cols_non_rep`
# - have *different* values in *all* columns `all_different_cols_non_rep`
# - have *different* values in *at least one* column of `any_different_cols_non_rep`
#
# Keep, both, (a, b) and (b, a)

any_different_cols_non_rep <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name")

all_same_cols_non_rep <-
  c("Metadata_cell_line", 
    "Metadata_Plate")

all_different_cols_non_rep <-
  c("Metadata_gene_name")

# ---- 5. Similarity to group ----

# Fetch similarities between
# a. all rows (except, optionally, `reference` rows)
# and
# b. all rows (except, optionally, `reference` rows)
#
# Do so for only those (a, b) pairs that
# - have *same* values in *all* columns of `all_same_cols_group`
# - have *different* values in *at least one* column of `any_different_cols_group`
#
# Keep, both, (a, b) and (b, a)

all_same_cols_group <-
  c("Metadata_cell_line",
    "Metadata_gene_name")

any_different_cols_group <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name")

# ---- 6. Combine 1-5 and annotate the similarity matrix ----

annotation_cols <-
  c("Metadata_cell_line",
    "Metadata_gene_name",
    "Metadata_pert_name")

munged_sim <-
  sim_munge(
    sim_df,
    metadata,
    reference,
    all_same_cols_rep = all_same_cols_rep,
    all_same_cols_rep_ref = all_same_cols_rep_ref,
    all_same_cols_ref = all_same_cols_ref,
    any_different_cols_non_rep = any_different_cols_non_rep,
    all_same_cols_non_rep = all_same_cols_non_rep,
    all_different_cols_non_rep = all_different_cols_non_rep,
    any_different_cols_group = any_different_cols_group,
    all_same_cols_group = all_same_cols_group,
    annotation_cols = annotation_cols,
    drop_group = drop_group
  )

```

# Check

Check for duplicates 

```{r}
stopifnot(
  munged_sim %>%
    anti_join(reference, by = colnames(reference)) %>%
    unite("id12", all_of(c("id1", "id2")), sep = ":") %>%
    pull("id12") %>%
    duplicated() %>%
    any() %>%
    not()
)

stopifnot(
  munged_sim %>%
    inner_join(reference, by = colnames(reference)) %>%
    filter(type == "rep") %>%
    unite("id12", all_of(c("id1", "id2")), sep = ":") %>%
    pull("id12") %>%
    duplicated() %>%
    any() %>%
    not()
)

stopifnot(
  munged_sim %>%
    inner_join(reference, by = colnames(reference)) %>%
    filter(type == "rep_group") %>%
    unite("id12", all_of(c("id1", "id2")), sep = ":") %>%
    pull("id12") %>%
    duplicated() %>%
    any() %>%
    not()
)

stopifnot(
  munged_sim %>%
    inner_join(reference, by = colnames(reference)) %>%
    filter(type != "rep" && type != "rep_group") %>%
    unite("id12", all_of(c("id1", "id2")), sep = ":") %>%
    pull("id12") %>%
    duplicated() %>%
    any() %>%
    not()
)
```

Report counts

```{r}
all_sim_counts <-
  munged_sim %>%
  group_by(across(all_of(c("id1", all_same_cols_rep, "type")))) %>%
  tally() %>%
  arrange(across(all_of(all_same_cols_rep))) %>%
  pivot_wider(names_from = "type",
              names_prefix = "n_",
              values_from = n,
              values_fill = 0)
```


Replicate counts (`n_pert` perturbations have `n_rep` replicate similarities)

```{r}
all_sim_counts %>%
  group_by(n_rep) %>%
  tally(name = "n_pert") %>%
  select(n_pert, n_rep)
```

Non-replicate counts (`n_pert` perturbations have `n_non_rep` non-replicate similarities)

```{r}
all_sim_counts %>%
  group_by(n_non_rep) %>%
  tally(name = "n_pert") %>%
  select(n_pert, n_non_rep)
```


Reference counts (`n_pert` perturbations have `n_ref` reference similarities)

```{r}
all_sim_counts %>%
  group_by(n_ref) %>%
  tally(name = "n_pert") %>%
  select(n_pert, n_ref)
```

Group counts (`n_pert` perturbations have `n_rep_group` reference similarities)

```{r}
all_sim_counts %>%
  group_by(n_rep_group) %>%
  tally(name = "n_pert") %>%
  select(n_pert, n_rep_group)
```

# Inspect 

```{r}
norm_non_rep_metrics <- 
  sim_metrics(munged_sim, "non_rep", calculate_grouped = TRUE)

norm_ref_metrics <- 
  sim_metrics(munged_sim, "ref", calculate_grouped = TRUE)

per_row_metrics <-
  inner_join(
    norm_non_rep_metrics[["per_row"]],
    norm_ref_metrics[["per_row"]],
    by = c("id1", all_same_cols_rep, "sim_mean_agg")
  )

per_set_metrics <-
  inner_join(
    norm_non_rep_metrics[["per_set"]],
    norm_ref_metrics[["per_set"]],
    by = c(all_same_cols_rep, "sim_mean_agg_mean_agg", "sim_mean_agg_median_agg")
  )

per_set_group_metrics <-
  inner_join(
    norm_non_rep_metrics[["per_set_group"]],
    norm_ref_metrics[["per_set_group"]],
    by = c(all_same_cols_rep, "sim_mean_agg", "sim_median_agg")
  )
```

  
```{r}
plot_scatter <- function(df, x0, y0) {
  df %>%
    ggpubr::ggscatter(x = x0,
                      y = y0) +
    ggpubr::stat_cor(method = "spearman") + coord_equal()
}
```


```{r}
plot_scatter(
  per_row_metrics, 
  "sim_scaled_mean_agg_non_rep", 
  "sim_mean_agg") + 
  ggtitle("Replicate level, wrt non_reps")

plot_scatter(
  per_set_metrics, 
  "sim_scaled_mean_agg_non_rep_mean_agg", 
  "sim_mean_agg_mean_agg") + 
  ggtitle("Perturbation level, wrt non_reps (mean)")

plot_scatter(
  per_set_metrics, 
  "sim_scaled_mean_agg_non_rep_median_agg", 
  "sim_mean_agg_median_agg") + 
  ggtitle("Perturbation level, wrt non_reps (median)")
```


```{r}
plot_scatter(
  per_row_metrics, 
  "sim_scaled_mean_agg_ref", 
  "sim_mean_agg") + 
  ggtitle("Replicate level, wrt reference")

plot_scatter(
  per_set_metrics, 
  "sim_scaled_mean_agg_ref_mean_agg", 
  "sim_mean_agg_mean_agg") + 
  ggtitle("Perturbation level, wrt reference (mean)")

plot_scatter(
  per_set_metrics,
  "sim_scaled_mean_agg_ref_median_agg",
  "sim_mean_agg_median_agg") + 
  ggtitle("Perturbation level, wrt reference (median)")
```


```{r}
plot_scatter(
  per_row_metrics, 
  "sim_scaled_mean_agg_ref", 
  "sim_scaled_mean_agg_non_rep") + 
  ggtitle("Replicate level, wrt reference, wrt non_reps")

plot_scatter(
  per_set_metrics, 
  "sim_scaled_mean_agg_ref_mean_agg", 
  "sim_scaled_mean_agg_non_rep_mean_agg") + 
  ggtitle("Perturbation level, wrt reference, wrt non_reps (mean)")

plot_scatter(
  per_set_metrics, 
  "sim_scaled_mean_agg_ref_median_agg", 
  "sim_scaled_mean_agg_non_rep_median_agg") + 
  ggtitle("Perturbation level, wrt reference, wrt non_reps (median)")

```
```{r}
plot_scatter(
  per_set_group_metrics, 
  "sim_scaled_mean_agg_ref", 
  "sim_mean_agg") + 
  ggtitle("Perturbation level, wrt ref (mean)")

plot_scatter(
  per_set_group_metrics, 
  "sim_scaled_median_agg_ref", 
  "sim_median_agg") + 
  ggtitle("Perturbation level, wrt ref (median)")
```

```{r message=FALSE}
gritpy_df <- 
  read_tsv("https://raw.githubusercontent.com/broadinstitute/grit-benchmark/31d9812e0c267d3535c85a72ef6bd0104d1a2130/1.calculate-metrics/cell-health/results/cell_health_grit.tsv") %>%
  filter(barcode_control == "cutting_control" & cor_method == "pearson" & grit_replicate_summary_method == "mean") %>%
  arrange(cell_line, perturbation, group) %>%
  anti_join(
    drop_group %>%
      rename(group = Metadata_gene_name)
  ) %>%
  select(
    cell_line, perturbation, group, grit
  )
```


```{r}
gritr_df <-
  per_set_group_metrics %>%
  select(
    perturbation = Metadata_pert_name,
    group = Metadata_gene_name,
    grit = sim_scaled_mean_agg_ref,
    cell_line = Metadata_cell_line
  ) %>%
  arrange(cell_line, perturbation, group)

grit_compare <-
  inner_join(
    gritpy_df,
    gritr_df,
    by = c("cell_line", "perturbation", "group"),
    suffix = c("_py", "_r")
  )

plot_scatter(grit_compare %>% na.omit(), "grit_py", "grit_r")
```

# Check one

```{r}
target_guide <- data.frame(
  Metadata_cell_line = "A549",
  Metadata_gene_name = "AKT1",
  Metadata_pert_name = "AKT1-1")

sister_guide <- data.frame(
  Metadata_cell_line = "A549",
  Metadata_gene_name = "AKT1",
  Metadata_pert_name = "AKT1-2")

cutting_controls <- data.frame(
  Metadata_cell_line = c("A549", "A549", "A549"),
  Metadata_gene_name = c("Chr2", "Luc", "LacZ")
)

metadata <- 
  profiles %>%
  select(matches("Metadata")) %>% 
  select(-Metadata_WellCol, -Metadata_WellRow)

features <- function(df) t(as.matrix(df %>% select(-matches("^Metadata_"))))

plate <- function(df) df$Metadata_Plate

p_target_guide <- profiles %>% inner_join(target_guide)

p_sister_guide <- profiles %>% inner_join(sister_guide)

p_cutting_controls <- profiles %>% inner_join(cutting_controls)

# get group similarities
s_group <- as.vector(cor(features(p_target_guide), features(p_sister_guide)))

# get control similarities (must be on same plate)
f_same_plate_target_v_control <- 
  outer(plate(p_target_guide), plate(p_cutting_controls), FUN = "==")

s_controls <-
  cor(features(p_target_guide), features(p_cutting_controls))[f_same_plate_target_v_control]

# check control counts
n_ref_count_test <- data.frame(n_ref = length(s_controls))

n_ref_count_check <- 
  all_sim_counts %>% 
  ungroup() %>% 
  inner_join(target_guide) %>% 
  summarise(n_ref = sum(n_ref))

stopifnot(n_ref_count_test == n_ref_count_check)


# check group counts
n_rep_group_count_test <- data.frame(n_rep_group = length(s_group))

n_rep_group_count_check <- 
  all_sim_counts %>% 
  ungroup() %>% 
  inner_join(target_guide) %>% 
  summarise(n_rep_group = sum(n_rep_group))

stopifnot(n_rep_group_count_test == n_rep_group_count_check)

s_group_check <- 
  munged_sim %>% 
  inner_join(target_guide) %>% 
  filter(type == "rep_group") %>% 
  pull("sim")

s_controls_check <- 
  munged_sim %>% 
  inner_join(target_guide) %>% 
  filter(type == "ref") %>% 
  pull("sim")

stopifnot(all.equal.numeric(sort(s_group_check), sort(s_group)))

stopifnot(all.equal.numeric(sort(s_controls_check), sort(s_controls)))
```


```{r}
n_ref_count_check
```


```{r}
n_rep_group_count_test
```
```{r}
data.frame(mean = mean(s_controls), sd = sd(s_controls))
```


```{r}
# compute grit
target_grit <- 
  target_guide %>%
  mutate(grit = mean(scale(s_group, mean(s_controls), sd(s_controls)))) %>%
  rename(
    perturbation = Metadata_pert_name,
    group = Metadata_gene_name,
    cell_line = Metadata_cell_line
  ) %>%
  select(perturbation, group, cell_line, grit)

target_grit
```


```{r}
target_grit_check <- 
gritr_df %>% inner_join(
  target_guide,
  by = c(
    "perturbation" = "Metadata_pert_name",
    "group" = "Metadata_gene_name",
    "cell_line" = "Metadata_cell_line"
  )
) %>%
  select(perturbation, group, cell_line, grit)


target_grit_check

```


```{r}
gritpy_df %>% inner_join(
  target_guide,
  by = c(
    "perturbation" = "Metadata_pert_name",
    "group" = "Metadata_gene_name",
    "cell_line" = "Metadata_cell_line"
  )
)
```

```{r}
# compute grit
target_group_cor <- mean(s_group)

data.frame(target_group_cor)
```


```{r}
per_set_group_metrics %>%
  inner_join(target_guide) %>%
  pivot_longer(-all_of(colnames(target_guide)))
```


